Advances in Artificial Intelligence for Data Visualization: Developing Computer Vision Models to Automate Reading of Data Plots, with Application to Predictive Model Diagnostics

### 2022-02-07 Thesis/01-chap1.Rmd
begin introduction

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, cache=TRUE)
# Load any R packages you need here
library(forecast)
library(ggplot2)
```

@@

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, cache=TRUE)
# Load any R packages you need here
library(ggplot2)
```

@@

## Artificial intelligence and predictive modelling

**Artificial intelligence** (AI) is the field of research concerned with understanding and building machines who can demonstrate intelligence. As discussed in @russell_artificial_2002, historically, there are disagreements among researchers on the definition of intelligence, which is caused by two critical questions:

1. Should AI act and think humanly or rationally? 
2. Without the thought process and reasoning, are behaviours sufficient to demonstrate intelligence?

Based on the answer to the above questions, four major approaches to pursue AI have been established. They can be summarized into a two by two table as shown in figure \@ref(fig:twodimai), where the row is "Human" vs "Rational", and the column is "Behaviour" vs "Thought". Positioning at the top right cell, the **rational agent approach** aims to build agent that perform mathematically perfect acts such that the best expected outcome can always be achieved. In contrast, the **"laws of thought" approach** focus on understanding the logic behind the rationality. Closely related to **cognitive science**, the **cognitive modelling approach** attempts to express theories of human cognition as computer program to mimic the thought process of human. Lastly, the **Turing test approach** is built upon the famous **Turing test** proposed by @turing_computing_1950. The test can be roughly described as, whether a human can distinguish another human from a computer with written communications only. To pass the test, several capabilities of computer are required. This includes **natural language processing**, **knowledge representation**, **automated reasoning** and **machine learning**. Some researchers argued that some degree of physical simulation of a person is still necessary, one such example is the **total Turing test** proposed by @harnad_other_1991. It adds new requirements to the list, including **computer vision**, **speech recognition** and **robotics** [@russell_artificial_2002].

Notably, all $6$ required capabilities have become major subfields of AI today.

Among The most influential one may be the machine learning, which highly rely on the predictive modelling framework.

Donoho (2017) in its summary of data science stated that the concept of this modelling culture could be traced back to an article written by Breiman (2001). In contrast to the generative modelling culture, which aims to develop stochastic models to make inferences about the data generating process, predictive modelling emphasizes the ability of the model to make accurate predictions.

Behind the success of AI and computer vision, a great proportion of their applications highly rely on the predictive modelling framework.

```{r twodimai, fig.cap="test"}
ggplot() +
  geom_hline(aes(yintercept = 1), col = "grey") +
  geom_vline(aes(xintercept = 1), col = "grey") +
  geom_text(aes(x = 0.5 , y = 1.5, label = "The Turing test approach")) +
  geom_text(aes(x = 1.5 , y = 1.5, label = "The rational agent approach")) +
  geom_text(aes(x = 0.5 , y = 0.5, label = "The cognitive modelling approach")) +
  geom_text(aes(x = 1.5 , y = 0.5, label = 'The "laws of thought" approach')) +
  theme_light() +
  xlab("") +
  ylab("") +
  scale_x_continuous(breaks = c(0.5, 1.5), labels = c("Human", "Rational"), limits = c(0, 2)) +
  scale_y_continuous(breaks = c(0.5, 1.5), labels = c("Thought", "Behaviour"), limits = c(0, 2)) +
  theme(panel.grid = element_blank(), axis.ticks = element_blank())
```

The rational agent approach

which impacts the version of AI pursued by researchers.

due to the definition of intelligence,

It has been pursued in many different ways by the researchers. The major separations between different approaches are the

reasoning and thoughts, focus on the results or the action. Act and think humanly or rationally.

Two dimensions

machine think humanly, think rationally, acting humanly, acting rationally

A rational agent is one that acts so as to achieve the best outcome or, when there is uncertainty, the best expected outcome.

versions due to

human performance

The cognitive modeling approach

The “laws of thought” approach

The rational agent approach

human vs. rational2 and thought vs. behavior

Some have defined intelligence in terms of fidelity to human performance, while others prefer an abstract, formal definition of intelligence called rationality—loosely speaking, doing the “right thing.” The subject matter itself also varies: some consider intelligence to be a property of internal thought processes and reasoning, while others focus on intelligent behavior, an external characterization.

As required by the Turing

As defined by Turing test,

The Turning test is well known

The field of artificial intelligence, or AI, is concerned with not just understanding but also building intelligent entities—machines that can compute how to act effectively and safely in a wide variety of novel situations.

## AI -> predictive modelling

## Predictive modelling -> Model diagnostics

## Limitation of MD/EDA (before 2009)

## Visual inference

## Limitation of visual inference

## Automatic visual inference -> Computer vision

### 2022-02-07 Thesis/_book/thesis.tex
begin introduction

@@

\hypertarget{visual-inference}{%
\subsection{Visual Inference}\label{visual-inference}}

@@

Thinking out of the box, \textcite{buja_statistical_2009} argued that there is actually no need for pre-specification of visual discoverable features. In model diagnostics, when the null hypothesis is rejected, the reasons for rejecting the hypothesis will also be known. This is because observers not only can point out the fact that visual discoveries have been found, but also can describe the particular features they observed. Those features will correspond to the subset of the collection of visual test statistics \(T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)\) which resulted in rejection.

@@

\hypertarget{type-i-error}{%
\subsubsection{Type-I error}\label{type-i-error}}

\hypertarget{type-ii-error}{%
\subsubsection{Type-II error}\label{type-ii-error}}

>>>>

\hypertarget{artificial-intelligence-and-predictive-modelling}{%
\section{Artificial intelligence and predictive modelling}\label{artificial-intelligence-and-predictive-modelling}}

\textbf{Artificial intelligence} (AI) is the field of research concerned with understanding and building machines who can demonstrate intelligence. As discussed in \textcite{russell_artificial_2002}, historically, there are disagreements among researchers on the definition of intelligence, which is caused by two critical questions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Should AI act and think humanly or rationally?
\item
  Without the thought process and reasoning, are behaviours sufficient to demonstrate intelligence?
\end{enumerate}

Based on the answer to the above questions, four major approaches to pursue AI have been established. They can be summarized into a two by two table as shown in figure \ref{fig:twodimai}, where the row is ``Human'' vs ``Rational'', and the column is ``Behaviour'' vs ``Thought''. Positioning at the top right cell, the \textbf{rational agent approach} aims to build agent that perform mathematically perfect acts such that the best expected outcome can always be achieved. In contrast, the \textbf{``laws of thought'' approach} focus on understanding the logic behind the rationality. Closely related to \textbf{cognitive science}, the \textbf{cognitive modelling approach} attempts to express theories of human cognition as computer program to mimic the thought process of human. Lastly, the \textbf{Turing test approach} is built upon the famous \textbf{Turing test} proposed by \textcite{turing_computing_1950}. The test can be roughly described as, whether a human can distinguish another human from a computer with written communications only. To pass the test, several capabilities of computer are required. This includes \textbf{natural language processing}, \textbf{knowledge representation}, \textbf{automated reasoning} and \textbf{machine learning}. Some researchers argued that some degree of physical simulation of a person is still necessary, one such example is the \textbf{total Turing test} proposed by \textcite{harnad_other_1991}. It adds new requirements to the list, including \textbf{computer vision}, \textbf{speech recognition} and \textbf{robotics} \autocite{russell_artificial_2002}. Notably, all \(6\) required capabilities have become major subfields of AI today.

\begin{figure}
\centering
\includegraphics{thesis_files/figure-latex/twodimai-1.pdf}
\caption{\label{fig:twodimai}test}
\end{figure}

The rational agent approach

which impacts the version of AI pursued by researchers.

due to the definition of intelligence,

It has been pursued in many different ways by the researchers. The major separations between different approaches are the

reasoning and thoughts, focus on the results or the action. Act and think humanly or rationally.

Two dimensions

machine think humanly, think rationally, acting humanly, acting rationally

A rational agent is one that acts so as to achieve the best outcome or, when there is uncertainty, the best expected outcome.

versions due to

human performance

The cognitive modeling approach

The ``laws of thought'' approach

The rational agent approach

human vs.~rational2 and thought vs.~behavior

Some have defined intelligence in terms of fidelity to human performance, while others prefer an abstract, formal definition of intelligence called rationality---loosely speaking, doing the ``right thing.'' The subject matter itself also varies: some consider intelligence to be a property of internal thought processes and reasoning, while others focus on intelligent behavior, an external characterization.

As required by the Turing

As defined by Turing test,

The Turning test is well known

The field of artificial intelligence, or AI, is concerned with not just understanding but also building intelligent entities---machines that can compute how to act effectively and safely in a wide variety of novel situations.

\hypertarget{ai---predictive-modelling}{%
\section{AI -\textgreater{} predictive modelling}\label{ai---predictive-modelling}}

\hypertarget{predictive-modelling---model-diagnostics}{%
\section{Predictive modelling -\textgreater{} Model diagnostics}\label{predictive-modelling---model-diagnostics}}

\hypertarget{limitation-of-mdeda-before-2009}{%
\section{Limitation of MD/EDA (before 2009)}\label{limitation-of-mdeda-before-2009}}

\hypertarget{visual-inference}{%
\section{Visual inference}\label{visual-inference}}

\hypertarget{limitation-of-visual-inference}{%
\section{Limitation of visual inference}\label{limitation-of-visual-inference}}

\hypertarget{automatic-visual-inference---computer-vision}{%
\section{Automatic visual inference -\textgreater{} Computer vision}\label{automatic-visual-inference---computer-vision}}

@@

\hypertarget{visual-inference-1}{%
\subsection{Visual Inference}\label{visual-inference-1}}

@@

Thinking out of the box, \textcite{buja_statistical_2009} argued that there is actually no need for pre-specification of visual discoverable features. In model diagnostics, when the null hypothesis is rejected, the reasons for rejecting the hypothesis will also be known. This is because observers can not only point out the fact that visual discoveries have been found, but also describe the particular visual features they observed. Those features will correspond to the subset of the collection of visual test statistics \(T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)\) which resulted in rejection. This argument helps justifies the validity of visual inference.

@@

With the validity of visual inference being justified, another aspect of hypothesis testing that needs to be addressed is the control of false positive rate or Type I error. Any visual statistic \(T^{(i)}(\boldsymbol{\mathrm{y}})\) needs to pair with a critical value \(c^{(i)}\) to form a hypothesis test. When a visual feature \(i\) is discovered by the observer from a plot, the corresponding visual statistic \(T^{(i)}(\boldsymbol{\mathrm{y}})\) may not be known as there is no general agreement on the measurement of the degree of presence of a visual feature. It is only the event that \(T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}\) is confirmed. Similarly, if any visual discovery is found by the observer, we say, there exists \(i \in I:~T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}\) \autocite{buja_statistical_2009}.

Using the above definition, the family-wise Type I error can be controlled if one can provide the collection of critical values \(c^{(i)}~(i \in I)\) such that \(P(\mathrm{there~exists~} i \in I: T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}|\boldsymbol{\mathrm{y}}) \leq \alpha\), where \(\alpha\) is the significance level. However, since the quantity of \(T^{(i)}(\boldsymbol{\mathrm{y}})\) may not be known, such collection of critical values can not be provided.

\textcite{buja_statistical_2009} proposed the lineup protocol as a visual test to calibrate the Type I error issue without the specification of \(c^{(i)}~(i \in I)\). It is inspired by the ``police lineup'' or ``identity parade'' which is the act of asking the eyewitness to identify criminal suspect from a group of irrelevant people. The protocol consists of \(m\) randomly placed data plots, where \(1\) plot is the actual data plot, and \(m-1\) null plots are produced by plotting data simulate from the null distribution which is consistent with the null hypothesis. Then, an observer who have not seen the actual data plot will be asked to point out the most different plot from the lineup.

Under the null hypothesis, it is expected that the actual data plot would have no distinguishable difference with the null plots, and the probability of the observer correctly picks the actual data plot is \(1/m\) due to randomness. If we reject the null hypothesis as the observer correctly picks the actual data plot, then the Type I error of this test is \(1/m\).

This provides us with an mechanism to control the Type I error, because \(m\) - the number of plots in a lineup can be chosen. Further, if we involve \(K\) independent observers in a visual test, and let \(X\) be a random variable denoting the number of observers correctly picking the actual data plot. Then, under the null hypothesis \(X \sim \mathrm{Binom}_{K,1/m}\), and therefore, the \(p\)-value of a lineup of size \(m\) evaluated by \(K\) observer is given as

\[P(X \geq x) = \sum_{i=x}^{K}{{K}\choose{i}}\left(\frac{1}{m}\right)^i\left(\frac{m-1}{m}\right)^{k-i},\]
where \(x\) is the realization of number of observers correctly picking the actual data plot \autocite{majumder_validation_2013}.
